Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-15 papers/ArticleZ80/source/paper.tex
ajustes

@@

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}
%
%
%
\author{Val\'{e}rio Medeiros Jr\inst{1}, David D\'{e}harbe\inst{2}}

@@

%\keywords{ Validation, Verification, Microcontroller, Hardware}
%
%\setcounter{secnumdepth}{1}
%\setcounter{chapter}{1}
%\setcounter{section}{1}

@@

The new model supports animation in ProB\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of microcontroller model. Such
library has many definitions of concepts common in microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using Z80 assembly language to analyse the code simulation
and the verification process.

The formal model of an instruction set has several
applications. Animation allows to simulate the execution of assembly
programs including support for interrupts, input and output
instructions. Other possible uses include documentation, the
construction of simulators, and can be possibly the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of such formalization
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers and build a reference model for a formal
compilation approach in B.

It is often the case that the official documentation of an instruction
uses textual description, mathematical formulas, examples,
etc. Sometimes, the instruction descriptions have a specific notation,
some informations are organized in different sections of document and
the textual descriptions are not standardized.  The manuals also have
semi-formal and informal descriptions, these descriptions may leave
errors, inconsistency and ambiguity.  For example, in the official
manual of the microcontroller Z80 \cite{Z80_manual}, the semantics of
an instruction (in particular its effects on the flag register) are
described informally, and is this description is scattered in
different pages.  Furthermore, the official manual of the
microcontroller Z80 has also several problems that were identified
over time~\cite{UndocumentedZ80}.

A solution to avoid ambiguity and inconsistencies is to specify
formally the assembly instruction set, making it possible to prove
properties of the instructions. The formal model also restricts the
definitions to correct typing and uses only well-defined
expressions. Furthermore, the developer may use the refinement
capabilities of the B method to specify at different levels of
abstraction and add more specific details or useful restrictions in
refined models.

@@

We have developed a reusable set of basic definitions to model
hardware concepts and data types concepts. The models contains
definitions to represent: registers, interruptions, input and
output ports, memory and instructions. These definitions are grouped
into separated development projects and are available as libraries.
Thus the workspace is composed of the following: 
\begin{description}

@@

The module also provides lemmas on negation that may be useful for the
users of the library to develop proofs:
$$
\begin{array}{l}
%  \textit{bit\_not}(0) = 1;  \textit{bit\_not}(1) = 0; \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$$
Conjunction is a binary function on bits and it is defined as:
$$
\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$$
The module provides the following lemmas for commutativity and associativity:
$$
\begin{array}{l}
%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
\end{array}
$$
The module provides definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar as for $\textit{bit\_and}$.

@@

Sequences are predefined objects in B and are natural candidate to
represent bit vectors. In B, sequences are functions whose the domain
is an integer range with lower bound 1 (one). Indices in bit vectors
usually range from 0 (zero) upwards and the model we propose obeys
this convention by making an offset where necessary, so that
predefined functions and proof rules may be used. We define bit
vectors as non-empty sequences of bits: $\textit{BIT\_VECTOR} =
\seq1(\textit{BIT})$. As $\textit{BIT\_VECTOR}$ is an infinite set,
this definition hinders the animation in ProB.  As a workaround, we
defined specialized types \textit{BYTE} and \textit{BV16} without
reference to \textit{BIT\_VECTOR}.

@@

We also define two functions $\textit{bv\_set}$ and
$\textit{bv\_clear}$ that, given a bit vector, and a position of the
bit vector, return the bit vector resulting from setting the
corresponding position to 1 and 0, and a function $\textit{bv\_get}$
that, given a bit vector and a valid position, returns the value of
the bit at that position. Only the first definition is shown here:
$$
\begin{array}{l}
\textit{bv\_set} \in \textit{BIT\_VECTOR} \times \nat \fun \textit{BIT\_VECTOR} \land \textit{bv\_set} =\\
\lambda v, n . (v \in \textit{BIT\_VECTOR} \land n \in \nat \land n <\textit{size}(v)
\mid v \lover \{ n+1 \mapsto 1 \})
\end{array}
$$

@@

\begin{table}
\caption{Details of basic data types}
\label{tab:types}
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
 \textit{Type\ Name} & \textit{Range} & \textit{Physical Size}  & \textit{Num. POs} & \textit{Library} \\\hline
 BIT & 0..1 & 1 bit &   118 & Hardware \\\hline %49 + 69 = 118 \\\hline
 BYTE & -- & 1 bytes &  153 & Hardware \\\hline%17 + 136 = 153\\\hline
 BV16 & -- & 2 bytes &  75 & Hardware \\ \hline%5 + 70 = 75 \\ \hline
 UCHAR & 0..255 & 1 byte &  30 & Types \\\hline %4 + 26 = 30 \\\hline
 SCHAR & -128..127 & 1 byte & 26 & Types \\\hline %2 + 24 = 26 \\\hline
 USHORT & 0..65.535 & 2 byte & 62 & Types \\\hline % 4 + 58 = 62 \\\hline
 SSHORT & -32.768..32.767 & 2 bytes & 58 & Types \\\hline %  2 + 56 = 58\\\hline
%  &  & Total & 522\\\hline %  2 + 56 = 58\\\hline
\end{tabular}
\end{center}
\end{table}

@@

Nearly 50\% of all these proof obligations shown in the
Table~\ref{tab:types} are verified automatically or with few user
commands. However many others proof obligations are difficult to
verify, as they involve functions that contain arithmetic expressions
that are difficult to manipulate with the interactive prover of
AtelierB.

\subsection{Verification} 
\label{sec:VerificationHardwareLibrary}

@@

To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause: this causes
corresponding proof obligations to be generated and verified. However,
we have not been able to build the corresponding proofs using the base
of rules included in the free distribution of AtelierB. So we
formalized new proof rules to check these properties in AtelierB and
used them in the interactive prover of AtelierB.

@@

The Z80 has 158 different instructions, including all the 78 from Intel 8080
microprocessor, and all of them were formally specified in B. These instructions are classified
into these categories: load and exchange; block transfer and search; arithmetic
and logical; rotate and shift; bit manipulation; jump, call and return;
input/output; and basic cpu control. Each category of instruction has different
elements of specification.

This section shows the elements that make up the different types of instructions.
The main elements are specified in the microcontroller module Z80 and parts of it
are presented below.  Basically, the state of microcontroller is formed by data 
from program counter, ports, registers, etc. The transitions between the states 
are actioned by execution of instruction or an external action. Groups of registers
are represented by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is
in the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined by the
clause \textit{OPERATIONS} and, in general, each instruction has three actions: update 
the program counter, update the flag register and its main effect.
%Several functions were created to Z80.
General functions that can be used in others microcontroller models are defined in
the modules of data types. Specific functions of the microcontroller are defined in the ALU (arithmetic logic unit) module.

@@

The main working register of Z80 is the accumulator ($\textit{rgs8(a0)}$) used for arithmetic/logic, input/output and loading/storing operations.
Another important element is the ``f'' register ($\textit{rgs8(f0)}$),  that is
used as a \textbf{flag register}. This register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning: $\textit{bv\_get(rgs8(f0),0)}$ - the carry bit; $\textit{bv\_get(rgs8(f0),1)}$ - the add/subtract bit; $\textit{bv\_get(rgs8(f0),2)}$ - the parity or overflow bit; $\textit{bv\_get(rgs8(f0),4)}$ - the half carry bit; 
$\textit{bv\_get(rgs8(f0),6)}$ - the zero bit; and $\textit{bv\_get(rgs8(f0),7)}$ - the sign bit.
%\begin{description}
%  \item[$\textit{bv\_get(rgs8(f0),0)}$] - The Carry bit
%  \item[$\textit{bv\_get(rgs8(f0),1)}$] - The Add/Subtract bit
%  \item[$\textit{bv\_get(rgs8(f0),2)}$] - The Parity or Overflow bit
%  \item[$\textit{bv\_get(rgs8(f0),4)}$] - The Half Carry bit
%  \item[$\textit{bv\_get(rgs8(f0),6)}$] - The Zero bit
%  \item[$\textit{bv\_get(rgs8(f0),7)}$] - The Sign bit 
%\end{description}
These bits also can be used to specify security properties for microcontroller programs.

@@

Another derived function is  $\textit{bv\_(ireg\_plus\_d)}$, this returns the value in the memory  address
returned by $\textit{bv\_ireg\_plus\_d}$ function and its definition is similar.

@@

In general, the instructions can access all memory address, but it is dangerous. The user may mistakenly
access and change data in memory. For added  security, it is important that the program instructions has
limited access by region. Thus the designer can specify address regions in a refinement model to restrict the access from
instructions. The address regions can be specified using constants
($\textit{PROGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$), these define respectively a range of restricted
address for: programs instructions, general purpose data and data stack.
%
%$
%\begin{array}{l}
%\textit{PROGRAM\_R\_ADR} = 0..16384 \land\\
%\textit{DATA\_R\_ADR} = 16385..49151 \land\\
%\textit{STACK\_R\_ADR} = 49152..65535
%\end{array}
%$

\textbf{Assuring the absence of overlapping of address regions:}
 \emph{To assure that address regions are well defined, then the designer must to verify the expression:}
%\begin{sloppypar}
%\hspace*{0.10in}
\it PROGRAM\_R\_ADR $\cap$ DATA\_R\_ADR $\cap$ $=$ \{\} $\land$ STACK\_R\_ADR $\cap$ DATA\_R\_ADR $\cap$ $=$ \{\} $\land$ PROGRAM\_R\_ADR $\cap$  STACK\_R\_ADR  $=$ \{\} \rm
%\end{sloppypar}

\textbf{Preserving the consistency of the memory:} \emph{In general, the access to
some address regions is dangerous. Then, each instruction has a specific pre-condition that verify
if the new address memory, that will be updated, is member of its region. For example, the $\textit{PUSH}$
program instruction allows write only in the region of stack ($\textit{STACK\_R\_ADR}$\rm).}

@@

Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has three categories of operations: a category represents the 
instructions of microcontrollers, a second category represents the input and output of data 
(shown in ~\ref{sec:modelingIO}) and the last category represents the external
actions (shown in~\ref{sec:externalactions}). A simple example of instruction category is
$\textit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The B tools does not allow
to use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module and it receives
a address memory and its new memory value. It also increments the program
counter ($\textit{pc}$) and update the refresh register ($\textit{r\_}$).
The other instructions have a similar structure.

@@

The $\textit{IN\_r(C)}$\footnoterecall{myfootnote} instruction is represented
by the following B operation. It receives an register identifier ``$\textit{rr}$'' and, it stores the value from ``rr'' in the $\textit{C}$ port address. Besides, it increments the program counter and updates the flag registers.

@@

The external actions change the state of the microcontroller, for example, refreshing the I/O ports
and interruptions request. The external actions are also modeled by operations and they are named with the
prefix ``$ext\_$'' and followed by the name of action. There are four external
actions: $ext\_update\_io\_ports$, $ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The
$ext\_update\_io\_ports$ just change the state of I/O port, see.

@@

The other external actions are related to interruptions. The interruptions allow
that the devices suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external
devices. When a routine is finished, then the CPU comes back to the last routine
that was interrupted.

@@

%\subsection{Considerations about the Z80 model}

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations guarantees: the correct use of data types, the developed
security properties  and the well-defined of all the expressions. %\footnote{An expression is called ``well-defined'' (or unambiguous) if its definition assigns it a unique interpretation or value.}.
Furthermore, the designer has a big flexibility to create new and specific security properties, this is very useful
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level. %There are many benefits when a formal model is
%developed, but to verify the model is not an easy task, the first version of Z80 model have already verified %completely,but the new Z80 model with support to animation in ProB has not been verified completely,
%because the changes in Z80 model are recent and the verification time is long.

>>>>

%TODO: Verificar possíveis diferenças do FORMATO LNCS

@@

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}

%
%
%
\author{Val\'{e}rio Medeiros Jr\inst{1}, David D\'{e}harbe\inst{2}}

@@

@@

The new model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of the microcontroller model. Such
a library has many definitions of concepts common to microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using the Z80 assembly language to analyse the code simulation
and the verification process.

The formal model of an instruction set has several
applications. Animation allows to simulate the execution of assembly
programs including support for interrupts, input and output
instructions. Other possible uses include documentation, the
construction of simulators, and can possibly be the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of such formalizations
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers and build a reference model for a formal
compilation approach in B.

It is often the case that the official documentation of an instruction
uses textual description, mathematical formulas, examples,
etc. Sometimes, the instruction descriptions have a specific notation,
some informations are organized in different sections of a document and
the textual descriptions are not standardized.  The manuals also have
semi-formal and informal descriptions, these descriptions may leave
errors, inconsistency and ambiguity.  For example, in the official
manual of the microcontroller Z80 \cite{Z80_manual}, the semantics of
an instruction (in particular its effects on the flag register) are
described informally, and this description is scattered in
different pages.  Furthermore, the official manual of the
microcontroller Z80 has also several problems that were identified
over time~\cite{UndocumentedZ80}.

A solution to avoid ambiguity and inconsistencies is to formally
specify the assembly instruction set, making it possible to prove
properties of the instructions. The formal model also restricts the
definitions to correct typing and uses only well-defined
expressions. Furthermore, the developer may use the refinement
capabilities of the B method to specify at different levels of
abstraction and add more specific details or useful restrictions in
refined models.

@@

We have developed a reusable set of basic definitions to model
hardware concepts and data types concepts. The models contain
definitions to represent: registers, interruptions, input and
output ports, memory and instructions. These definitions are grouped
into separated development projects and are available as libraries.
Thus the workspace is composed of the following: 
\begin{description}

@@

The module also provides lemmas on negation that may be useful for the
users of the library to develop proofs:
$$
\begin{array}{l}
%  \textit{bit\_not}(0) = 1;  \textit{bit\_not}(1) = 0; \\
\forall (\textit{bb}).(\textit{bb} \in \textit{BIT} \implies \textit{bit\_not}(\textit{bit\_not}(\textit{bb})) = \textit{bb})
\end{array}
$$
Conjunction is a binary function on bits and it is defined as:
$$
\begin{array}{l}
\textit{bit\_and} \in \textit{BIT} \times \textit{BIT} \fun \textit{BIT} \land \\
\forall (\textit{b1}, \textit{b2}).(\textit{b1}  \in \textit{BIT}  \land \textit{b2} \in \textit{BIT} \implies \\
\quad ((\textit{bit\_and}(\textit{b1}, \textit{b2}) = 1) \iff (\textit{b1} = 1)  \land  (\textit{b2} = 1)))
\end{array}
$$
The module provides the following lemmas for commutativity and associativity:
$$
\begin{array}{l}
%  \textit{bit\_and}(0,0) = 0;  \textit{bit\_and}(0,1) = 0; \\
%  \textit{bit\_and}(1,0) = 0;  \textit{bit\_and}(1,1) = 1; \\
\forall (\textit{b1},\textit{b2}).(\textit{b1} \in \textit{BIT} \land \textit{b2} \in \textit{BIT} \implies \\
\quad (\textit{bit\_and}(\textit{b1}, \textit{b2}) = \textit{bit\_and}(\textit{b2},\textit{b1})))\land \\
\forall (\textit{b1},\textit{b2},\textit{b3}).(\textit{b1} \in \textit{BIT} \land  \textit{b2} \in \textit{BIT} \land \textit{b3} \in \textit{BIT} \implies \\
\quad (\textit{bit\_and}(\textit{b1}, \textit{bit\_and}(\textit{b2},\textit{b3})) = \textit{bit\_and}(\textit{bit\_and}(\textit{b1},\textit{b2}),\textit{b3})))\\
% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 1) = \textit{b1})); \\
% \forall (\textit{b1}).(\textit{b1} \in \textit{BIT} \implies (\textit{bit\_and}(\textit{b1}, 0) = 0));
\end{array}
$$
The module provides definitions of $\textit{bit\_or}$ (disjunction)
and $\textit{bit\_xor}$ (exclusive disjunction), as well as lemmas on
those operators. These are standard and their expression in B is
similar to $\textit{bit\_and}$.

@@

Sequences are predefined objects in B and are natural candidates to
represent bit vectors. In B, sequences are functions whose domain
is an integer range with lower bound 1 (one). Indices in bit vectors
usually range from 0 (zero) upwards and the model we propose obeys
this convention by making an offset where necessary, so that
predefined functions and proof rules may be used. We define bit
vectors as non-empty sequences of bits: $\textit{BIT\_VECTOR} =
\seq1 (\textit{BIT})$. As $\textit{BIT\_VECTOR}$ is an infinite set,
this definition hinders the animation in ProB.  As a workaround, we
defined specialized types \textit{BYTE} and \textit{BV16} without
reference to \textit{BIT\_VECTOR}.

@@

We also define two functions $\textit{bv\_set}$ and
$\textit{bv\_clear}$ that, given a bit vector, and a position in a
bit vector, returns the bit vector resulting from setting the
corresponding position to 1 and 0, and a function $\textit{bv\_get}$
that, given a bit vector and a valid position, returns the value of
the bit at that position. Only the first definition is shown here:
$$
\begin{array}{l}
\textit{bv\_set} \in \textit{BIT\_VECTOR} \times \nat \fun \textit{BIT\_VECTOR} \land \\ \textit{bv\_set} =
\lambda v, n . (v \in \textit{BIT\_VECTOR} \land n \in \nat \land n <\textit{size}(v)
\mid v \lover \{ n+1 \mapsto 1 \})
\end{array}
$$

@@

\begin{table}[h]
\caption{Details of basic data types}
\label{tab:types}
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
 \textit{Type\ Name} & \textit{Range} & \textit{Physical Size}  & \textit{Num. POs} & \textit{Library} \\\hline
 BIT & 0..1 & 1 bit &   118 & Hardware \\\hline 
 BYTE & -- & 1 bytes &  154 & Hardware \\\hline
 BV16 & -- & 2 bytes &  75 & Hardware \\ \hline
 UCHAR & 0..255 & 1 byte &  30 & Types \\\hline
 SCHAR & -128..127 & 1 byte & 30 & Types \\\hline 
 USHORT & 0..65535 & 2 byte & 62 & Types \\\hline
 SSHORT & -32768..32767 & 2 bytes & 62 & Types \\\hline 
\end{tabular}
\end{center}
\end{table}

@@

Nearly 50\% of all these proof obligations shown in the
Table~\ref{tab:types} are verified automatically or with few proof
commands\footnote{The proof commands are steps that help the prover to find the proof.} .
However many others proof obligations are difficult to verify, as they involve functions
that contain arithmetic expressions that are difficult to manipulate with the interactive
prover of AtelierB. However, all theses 531 proof obligations were verified completely.

\subsection{Verification of hardware library and types library} 
\label{sec:VerificationHardwareLibrary}

@@

To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause: this causes
corresponding proof obligations to be generated and verified. However,
we have not been able to build the corresponding proofs using the 
rule base included in the free distribution of AtelierB. So we
formalized new proof rules to check these properties in AtelierB and
used them in the interactive prover of AtelierB.

@@

The Z80 has 158 different instructions, including all the 78 from the Intel 8080
microprocessor, and all of them were formally specified in B. These instructions are classified
into these categories: load and exchange; block transfer and search; arithmetic
and logical; rotate and shift; bit manipulation; jump, call and return;
input/output; and basic cpu control. Each category of instruction has different
elements of specification.

This section shows the elements that make up the different types of instructions.
The main elements are specified in the microcontroller module Z80 and parts of it
are presented below.  Basically, the state of microcontroller is formed by data 
from program counter, ports, registers, etc. The transitions between the states 
are actioned by execution of instruction or an external action. Groups of registers
are represented by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is
in the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined by the
clause \textit{OPERATIONS} and, in general, each instruction has three actions: update 
the program counter, update the flag register and its main effect.
%Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined in
the modules of data types. Specific functions of the microcontroller are defined in the ALU (arithmetic logic unit) module.

@@

The main working register of Z80 is the accumulator ($\textit{rgs8(a0)}$) used for arithmetic/logic, input/output and loading/storing operations.
Another important element is the ``f'' register ($\textit{rgs8(f0)}$),  that is
used as a \textbf{flag register}. This register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning: %$\textit{bv\_get(rgs8(f0),0)}$ - the carry bit; $\textit{bv\_get(rgs8(f0),1)}$ - the add/subtract bit; $\textit{bv\_get(rgs8(f0),2)}$ - the parity or overflow bit; $\textit{bv\_get(rgs8(f0),4)}$ - the half carry bit; $\textit{bv\_get(rgs8(f0),6)}$ - the zero bit; and $\textit{bv\_get(rgs8(f0),7)}$ - the sign bit.
\begin{description}
  \item[$\textit{bv\_get(rgs8(f0),0)}$] - The Carry bit.
  \item[$\textit{bv\_get(rgs8(f0),1)}$] - The Add/Subtract bit.
  \item[$\textit{bv\_get(rgs8(f0),2)}$] - The Parity or Overflow bit.
  \item[$\textit{bv\_get(rgs8(f0),4)}$] - The Half Carry bit.
  \item[$\textit{bv\_get(rgs8(f0),6)}$] - The Zero bit.
  \item[$\textit{bv\_get(rgs8(f0),7)}$] - The Sign bit.
\end{description}
These bits can also be used to specify security properties for microcontroller programs.

@@

Another derived function is  $\textit{bv\_(ireg\_plus\_d)}$, which returns the value in the memory  address
returned by the $\textit{bv\_ireg\_plus\_d}$ function and its definition is similar.

@@

In general, the instructions can access all memory addresses, but it is dangerous. The user may mistakenly
access and change data in memory. For added  security, it is important that the program instructions has
limited access by region. Thus the designer can specify address regions in a refinement model to restrict the access from
instructions. The address regions can be specified using constants
($\textit{PROGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$), these define respectively a range of restricted
address for: programs instructions, general purpose data and data stack.
%
%$
%\begin{array}{l}
%\textit{PROGRAM\_R\_ADR} = 0..16384 \land\\
%\textit{DATA\_R\_ADR} = 16385..49151 \land\\
%\textit{STACK\_R\_ADR} = 49152..65535
%\end{array}
%$

\textbf{Assuring the absence of overlapping of address regions:}
 \emph{To assure that address regions are well defined, then the designer must verify the expression:}
%\begin{sloppypar}
%\hspace*{0.10in}
\it PROGRAM\_R\_ADR $\cap$ DATA\_R\_ADR $\cap$ $=$ \{\} $\land$ STACK\_R\_ADR $\cap$ DATA\_R\_ADR $\cap$ $=$ \{\} $\land$ PROGRAM\_R\_ADR $\cap$  STACK\_R\_ADR  $=$ \{\} \rm
%\end{sloppypar}

\textbf{Preserving the memory consistency:} \emph{In general, the access to
some address regions is dangerous. Then, each instruction has a specific pre-condition that verifies
if the new address memory, that will be updated, is a member of its region. For example, the $\textit{PUSH}$
program instruction permits write only in the region of stack ($\textit{STACK\_R\_ADR}$\rm).}

@@

Each instruction is represented by a B operation in Z80 module. The main
module (\textit{Z80}) has three categories of operations: a category represents the 
instructions of microcontrollers, a second category represents the input and output of data 
(shown in ~\ref{sec:modelingIO}) and the last category represents the external
actions (shown in~\ref{sec:externalactions}). A simple example of instruction is
$\textit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The B tools does not allow
to use parentheses in identifiers, and the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the actual specification.} shown below. The
pre-defined functions are necessary many times to model the instructions, these
functions facilitate the construction of instruction set model. By default, all
parameters in operations are either predefined elements in the model or
integers values in decimal representation. This instruction use the
$\textit{updateAddressMem}$ operation from \textit{Memory} module and it receives
a address memory and its new memory value. It also increments the program
counter ($\textit{pc}$) and update the refresh register ($\textit{r\_}$).
The other instructions have a similar structure.

@@

The $\textit{IN\_r(C)}$\footnoterecall{myfootnote} instruction is represented
by the following B operation. It receives a register identifier ``$\textit{rr}$'' and, it stores the value from ``rr'' in the $\textit{C}$ port address. Besides, it increments the program counter and updates the flag registers.

@@

The external actions change the state of the microcontroller, for example, refreshing the I/O ports
and interruption request. The external actions are also modeled by operations and they are named with the
prefix ``$ext\_$'' and followed by the name of action. There are four external
actions: $ext\_update\_io\_ports$, $ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The instruction
$ext\_update\_io\_ports$ just changes the state of I/O port, see.

@@

The other external actions are related to interruptions. Interruptions allow
devices to suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external
devices. When a routine is finished, then the CPU comes back to the last routine
that was interrupted.

@@

\subsection{Verification process}% ESTOU TRABALHANDO AQUI

% parágrafo inicial
This project has many proof obligations and its construction of then model was a difficult task, as several
% iterations were needed to provide the good library definitions
% as well as to optimize the definition of the functionality of
% the microprocessor instructions by grouping common data
% manipulation into auxiliary functions.

% [O que as provas garantem? citar que dificeis, que é necessário buscar técnicas eficientes
% para realizar as provas ]

The proof obligations allow to verify the data types, important system
properties and if the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees, because they can set many
safety rules. However, the model can be very difficult to prove.

% Conclusão: toque final
Several iterations were needed to provide the good library definitions as well as
to fine-tune the model of the microcontroller instructions by factoring common
functionalities into auxiliary definitions.

%OK -> como as regras foram utilizadas ?  Respondido na seção das regras -

%-> Quanto tempo foi gasto nessa verificação ? a minha dissertação diz que aproximadamente dois meses. Utilizando as regras este tempo caiu p/ 5 dias (linkar com este parágrafo)

%-> Aproximadamente quanto tempo de processamento elas reduziram ?
% Porque o modelo não foi verificado completamente ? Isto é viavel ?
 %There are many benefits when a formal model is
%developed, but to verify the model is not an easy task, the first version of Z80 model have already verified %completely,but the new Z80 model with support to animation in ProB has not been verified completely,
%because the changes in Z80 model are recent and the verification time is long.

TODO:ENGLISH REVIEW

%Decomposição do módulo Z80
The Z80 module was decomposed in 8 modules of instructions, it was very helpful to decrease the time verification, because the AtelierB can start the simultaneously 8 instances of provers reducing the time verification by up to 8 times, when it uses a processor with 8 cores.

 %-> O que significa cada coluna e linha da tabela ?  	
The follow table~\ref{tab:completestatistcs} contains the results of verification process, it shows only the number of no obvious proof obligations and the titles of columns use acronyms: {\small \textbf{PO}} - number of proof obligations; {\small \textbf{UPO}} - number of unproved proof obligations; {\small \textbf{WD PO}} - number of proof obligations  related to well definition lemmas; {\small \textbf{UWD PO}} - number of unproved proof obligations related to well definition lemmas.

\begin{table} [h]
\caption{Verification statistics }
\label{tab:completestatistcs}
\begin{center} 
\begin{tabular}{|p{3.1cm}|c|c||c|c||c|c|c|}
\hline 
{\small \textbf{Component}}&{\small \textbf{PO}}&{\small \textbf{UPO}}&{\small \textbf{WD PO}}&{\small \textbf{UWD PO }}&{\small \textbf{Total PO}}&{\small \textbf{Total UPO}}& {\small \textbf{Total Percentage}}  \\\hline
BIT&49&0&69&0&118&0&100\%\\\hline
BV16&6&0&69&0&75&0&100\%\\\hline
BYTE&18&0&136&0&154&0&100\%\\\hline
POWER&3&0&4&0&7&0&100\%\\\hline
POWER2&18&0 &0&0&18&0&100\%\\\hline
SCHAR&4&0&26&0&30&0&100\%\\\hline
SSHORT&4&0&58&0&62&0&100\%\\\hline
TYPES&71&0&0&0&71&0&100\%\\\hline
UCHAR&4&0&26&0&30&0&100\%\\\hline
USHORT&4&0&58&0&62&0&100\%\\\hline
ALU&51&22&117&47&168&69&59\%\\\hline
MEMORY&13&0&0&0&13&0&100\%\\\hline
Z80 -  {\scriptsize Initialization, Properties, Assertions} &66&9&150&11&216&20&91\%\\\hline
Z80 - {\scriptsize Input and Output}&46&11&281&16&327&27&92\%\\\hline
Z80 - {\scriptsize Logic Arithmetic 1}&88&33&343&84&431&117&73\%\\\hline
Z80 - {\scriptsize Logic Arithmetic 2}&40&20&178&27&218&47&79\%\\\hline
Z80 - {\scriptsize Bit Control}&111&34&763&157&874&191&78\%\\\hline
Z80 - {\scriptsize External Actions}&31&0&60&38&91&38&59\%\\\hline
Z80 - {\scriptsize General 1}&81&19&420&20&501&39&92\%\\\hline
Z80 - {\scriptsize General 2}&26&10&139&14&165&24&86\%\\\hline

\textbf{Total}&\textbf{734}&\textbf{158}&\textbf{2897}&\textbf{364}&\textbf{3631}&\textbf{569}&\textbf{84}\%\\\hline

\end{tabular}
\end{center} 
\end{table}

% Resumo de como aconteceu o processo de verificação, (melhor está após as estátisticas) ( )Atualizar dados
However, few proof commands need to be used to prove most proof obligations. As there are many similar assembly instructions,
some human-directed proofs, when replayed, could discharge other proof
obligations. A good example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of WD proofs. We also set up a proving environment
consisting of networked computers to take advantage of the distribution
facilities now provided in the B development environment. Finally, all of the
2926 proof obligations were proved using the tool support of the development
environment.

\subsection{Considerations about the Z80 model}
